---
title: "Basketball_Spending"
format: html
editor: visual
embed-resources: true
---

```{r}
library(dplyr)
library(tidyverse)
library(readr)
 library(stringr)
library(fuzzyjoin)
library(plotly)
library(shiny)

sports <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-03-29/sports.csv')

basketball_stat <- readr::read_csv("data/College Basketball Dataset-2019.csv")

basketball <- readr::read_csv("data/College Basketball Dataset.csv")

basketball <- basketball |> filter(YEAR %in% c("2015", "2016", "2017", "2018"))

```

```{r}
basketball_2015_2019 <- bind_rows(basketball, basketball_stat) |>

  mutate(YEAR = ifelse(is.na(YEAR), "2019", YEAR)) |>
  select(TEAM, CONF, G, W, POSTSEASON, SEED, YEAR) |>
  mutate(TEAM = TEAM |>
  str_to_lower())


  basketball_2015_2019$YEAR <- as.numeric(basketball_2015_2019$YEAR)

basketball_2015_2019
```

```{r}
basketball_money <- sports |> filter(sports == "Basketball") |>
  filter(classification_code == "1")|>
  select(year, unitid, institution_name, state_cd, classification_code, classification_name, sector_cd, sector_name, partic_men, partic_women, sum_partic_men, sum_partic_women, rev_men, rev_women, total_rev_menwomen, exp_men, exp_women, total_exp_menwomen, sports) |>
  rename(YEAR = year) |>
  rename(TEAM = institution_name) |>
  mutate(TEAM = TEAM |>
           str_to_lower() |>
           str_replace_all("university", "") |>
           str_replace_all("college", "") |>
           str_replace_all("campus", "") |>
           str_replace_all("of ", "") |>
           str_replace_all("at ", "") |>
           str_replace_all("main ","") |>
           str_replace_all("station ", "") |>
           str_replace_all("-", " ") |>
           str_replace_all("chapel hill", "") |>
           str_replace_all("ann arbor", "" ) |>
           str_replace_all("madison", "") |>
           str_replace_all("raleigh", "" ) |>
           str_squish())
```

```{r}
df <- stringdist_left_join(
  basketball_2015_2019, basketball_money,
  by = "TEAM",
  method = "osa",           # Optimal String Alignment
  max_dist = 1,             # Allow up to 2 changes (add, drop, swap)
  distance_col = "str_dist"
)

df <- df |> filter(YEAR.x == YEAR.y) |>
  rename(YEAR = YEAR.x,
         TEAM = TEAM.x) |>
  select(- YEAR.y,
         - TEAM.y) |>
  mutate(in_tournament = !is.na(POSTSEASON)) |>
mutate(profit_men = rev_men - exp_men)
```

Abstract:

This project analyzes a data set of college basketball teams, focusing on the relationship between financial performance (such as revenue, expenses, and profit) and on-court success, including wins and postseason outcomes. Using interactive visualizations built in a Shiny app, the analysis allows users to explore distributions and relationships across different conferences and years. Key findings suggest that while teams with higher profits and revenues often tend to perform better and advance further in the postseason, the relationship is not perfectly linear, indicating that financial resources alone do not guarantee success. Other factors, such as team performance metrics, also play a significant role in determining postseason advancement. Some teams during this period displayed large profit numbers, with less success (Louisville). While if you find any cinderella run in March Madness, their profit margin is extremely low compared to any power 5 team.

Introduction:

This project examines the relationship between financial performance and competitive success in Division I college basketball. The data set includes team-level information across multiple years (2015-2019), with variables such as total wins, games played, revenue, expenses, and calculated profit, along with postseason outcomes indicating how far each team advanced in the NCAA tournament. By combining financial and performance metrics, the data provide an opportunity to explore whether more financially successful programs also achieve greater success on the court.

The primary questions of interest in this analysis are: How are financial variables such as revenue and profit distributed across teams and conferences? Is there a relationship between a team’s financial resources and its regular season success, measured by wins? Do teams with higher profits or revenues tend to advance further in the postseason? and how do these relationships vary across conferences and years? Through interactive visualizations, this project aims to uncover patterns and assess whether financial strength is a meaningful predictor of basketball performance, while also highlighting the limitations of such a relationship.

Comparing teams who made the tournament versus those who did not

```{r}
tourney_vs_no_tourney <- df |> group_by(in_tournament)

tourney_plot <- ggplot(tourney_vs_no_tourney, aes(x = YEAR, y = profit_men,  color = in_tournament,
                                  label = TEAM)) +
  geom_point() +
  geom_smooth(se = FALSE) +
    scale_y_continuous(labels = scales::comma) +
  labs(title = "Analyzing Teams in the Tournament vs not in Tournament"  ,
       y = "profit")
```

```{r}
ggplotly(tourney_plot, tooltip = "label")
```

The visual above depicts teams who made the NCAA March Madness tournament from 2015 to 2019 with profit of each team during each year. Based on the visual, there is clear correlation between teams who made the tournament versus those who did not. Teams who made the tournament were in the range of 5 million dollars profit each year. While teams who did not make the tournament were in the realm of 500,000 dollars to maybe $1.5 million.

Some things that stand out in this image is that there are some very large profits from teams that could easily skew the data. There is most likely enough data points to keep the data as a good visual but there are some teams that have much larger profits than others. Each year, Louisville had the highest profit and it wasn't even close with over 20 million dollars in profit. This is the aftermath of the 2013 season where Louisville won the national title. It did not go in there favor, as two of these five years they did not make the tournament. While a few years following this, Rick Pitino (Louisville coach) was found to be paying players and was fired. Their national championship was vacated and Louisville has not been the same team since in the ACC.

The same can be seen at the bottom of visual. If you hover over the very bottom team in each year, it is Notre Dame. They are losing money in every year and also missed the tournament two of the five years. This is interesting as they are also a power 5 team, showing that the majority of mid major schools operate in a more reserved fashion to ensure not losing money because they cannot afford to. While somebody like Notre Dame can afford to lose money on something like basketball because they are a prestigious university who definitely make large sums of money in other ways. They are a very big football school who definitely make enough money to make up for basketball losses.

College basketball during this period of time was known for a variety of teams paying players which ultimately led to the implementation of NIL. This visual, although difficult to see individual teams gives a great insight into the idea that making more money as a basketball team, in turn, leads to a better chance of making the NCAA tournament.

Looking at the Blue Bloods of college basketball

```{r}
blue_bloods <- df |>
 filter(TEAM %in% c("duke", "north carolina", "kentucky", "kansas", "villanova", "connecticut", "ucla")) |>
  group_by(TEAM)

blue_bloods_plot <- ggplot(data = blue_bloods, aes(x = YEAR, y = profit_men, color = TEAM,
                               label = POSTSEASON)) +
  geom_point() +
   geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  scale_color_viridis_d() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Profits and Finishes of Blue Blood Teams",
       y = "profit")
```

```{r}
ggplotly(blue_bloods_plot, tooltip = "label")
```

The term blue bloods has long been regarded as the best teams in college basketball. It is an ongoing thing in basketball that the best teams have majority blue in their logo's. These teams make up the majority of winners in the March Madness tournament. There has been only 1 team in the 21st century without blue in their logo to win March Madness. With that being Baylor following this data in 2021. During these years, all five teams to win March Madness had primarily blue in their logo. In 2015, Duke won, 2016, Villanova, 2017, North Carolina, 2018, Villanova, and finally 2019 was Virginia. Virginia being the only school without being primarily blue in its logo.

The teams I looked at in this visual are UCONN, Duke, Kansas, Kentucky, and North Carolina. Each of these teams generally make deep runs in the NCAA tournament and make a lot of money from basketball as well. All of these schools primary sport is basketball. What can be seen here is that UCONN is the worst team in this five year stretch by far with only one tournament birth. They had declining revenue in each season as the team was not doing well. Following these years, UCONN has come back to the top, winning 2 national championships in the last five years. North Carolina and Kentucky had a clear upward trend in profits over this period of time. This is most likely due to their postseason success leading to more fan engagement in the following years. While Kansas and Duke had consistent profits each year and consistent postseason success as well. During 2015, Duke had the highest profit at nearly 15 million dollars compared to UCONN, with roughly 2.5 million dollars. Compared to 2019, when North Carolina led the group profits with roughly 17 million dollars and UCONN was at the bottom who actually lost money. This makes sense as this was the first season under a new coach (Dan Hurley). He then turned the team around and UCONN is successful again.

Profits versus Postseason success

```{r}
df <-  df |> mutate(POSTSEASON = as.numeric(factor(POSTSEASON,
    levels = c("R68", "R64", "R32", 
               "S16", "E8", "F4", "2ND" ,"Champions")
  )))
```

```{r}
finish_vs_profit <- ggplot(df, aes(x = profit_men, y = POSTSEASON, color = W,
                                   label = TEAM)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Team Profit vs. Postseason Success",
    x = "Profit",
    y = "Postseason Round"
  ) +
    scale_color_viridis_c() +
  scale_x_continuous(labels = scales::comma)
```

```{r}
ggplotly(finish_vs_profit, tooltip = "label")
```

  So far, we have looked directly into postseason success based on each teams profits. While another clear indicator of postseason success is the amount of wins each team as in the regular season. We all know there is a relationship between regular season wins and the eventual finish in March Madness. As the 1 seed in the tournament, who are the 4 best regular season teams in the country have made up 18 of 25 national champions (2000-2025). This shows that teams who have more regular season wins tend to do much better in the tournament. 
  
  In this visual, a postseason round of 2 means that you lost in the round of 64 (the first round). There are a few teams below 2, they were in the play in games and lost during those (round of 68). With round 8 meaning you won the national championship. In this model, you can clearly see the trend of more profit equals more postseason success by getting to farther rounds. Within this, the teams with more wins in the regular season also very evident in getting farther in the postseason. It is a bit varied in the beginning, with a good amount of variation in rounds 3 or below. This shows that plenty of teams with less wins also won a round or two in the tournament. However, teams who have large success and win a lot of games all have a lot wins. Most teams getting to the elite 8 have over 30 wins on the season. Syracuse is one outlier here, winning in the range of 25 regular season games, while making it all the way to the final 4. 
  
  This visual does a good job showing a different view from just direct profit versus postseason success. Although there is definitely evidence of this. Regular season wins have a huge effect as well. There have been only 5 teams in the 21st century to win the tournament as 3 seed or lower. Three of those being three seeds, one four seed, and one seven seed. To win March Madness, the overwhelming majority says that you must be high seed to achieve great success in the tournament. 

Shiny app for Teams

```{r}
df_one_conf <- df |> filter(CONF == "ACC") 

ggplot(df_one_conf, aes(x = W)) +
  geom_histogram()
```

```{r}
ui <- fluidPage(

  titlePanel("College Basketball Model"),
  sidebarLayout(
    sidebarPanel(
      selectInput("year_choose",
                  "Choose a Year:",
                  choices = unique(df$YEAR)),

      selectInput("conf_choose",
                  "Choose Conference:",
                  choices = unique(df$CONF)),

      selectInput("variable_choose",
                  "Histogram Variable:",
                  choices = c("W","G","profit_men","rev_men","exp_men","POSTSEASON", "in_tournament")),

      sliderInput("n_bins",
                  "Number of Bins:",
                  min = 5, max = 30, value = 15),

      selectInput("x_var",
                  "X Variable (Relationship Plot):",
                  choices = c("profit_men","rev_men","exp_men","W","G")),

      selectInput("y_var",
                  "Y Variable:",
                  choices = c("W","POSTSEASON","profit_men","rev_men"))
    ),

    mainPanel(
      plotOutput("hist_conference"),
      plotly::plotlyOutput("scatter_plot")
    )
  )
)

server <- function(input, output, session) {

  data <- reactive({
    df |> filter(
      YEAR == input$year_choose,
      CONF == input$conf_choose
    )
  })

  output$hist_conference <- renderPlot({

  df_one_conf <- data()

  variable <- df_one_conf[[input$variable_choose]]

  if (is.numeric(var)) {

    ggplot(df_one_conf, aes(x = variable)) +
      geom_histogram(fill = "lightblue", bins = input$n_bins) +
      labs(title = glue::glue(input$variable_choose, " Distribution")) +
      theme_minimal(base_size = 16) +
      scale_x_continuous(labels = scales::comma)

  } else {

    ggplot(df_one_conf, aes(x = variable)) +
      geom_bar(fill = "lightblue") +
      labs(title = glue::glue(input$variable_choose, " Distribution")) +
      theme_minimal(base_size = 16) +
      scale_x_continuous(labels = scales::comma)
  }
})
  
  output$scatter_plot <- plotly::renderPlotly({

  df_one_conf <- data()

  scatterplot <- ggplot(df_one_conf, aes(x = .data[[input$x_var]], y = .data[[input$y_var]],
    text = paste("Team:", TEAM,
                 "<br>Year:", YEAR,
                 "<br>Conf:", CONF)
  )) +
    geom_point(alpha = 0.7) +
    scale_x_continuous(labels = scales::comma) +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = glue::glue(input$y_var, " vs ", input$x_var),
      x = input$x_var,
      y = input$y_var
    ) +
    theme_minimal()

  plotly::ggplotly(scatterplot, tooltip = "text")
})
}
  shinyApp(ui, server)
```


```{r}
ggplot(df, aes(x = W)) +
      geom_histogram(fill = "lightblue")  +
      theme_minimal(base_size = 16) +
  labs(title = "Looking at Wins in Data Set")

ggplot(df, aes(x = W, y = profit_men)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Comparing Wins vs Profit") +
   scale_y_continuous(labels = scales::comma)

ggplot(df, aes(x = profit_men)) +
  geom_histogram(fill = "lightblue")  +
      theme_minimal(base_size = 16) +
  labs(title = "Looking at Profit in Data Set") +
     scale_x_continuous(labels = scales::comma)

```


The shiny App allows for a clear interactive way for somebody to analyze the effects of the variables on basketball success. Although I did make a large amount of visuals, it is always good for somebody to be able to look into certain things that interest them. The histogram allows you to easily see things like wins, revenue, and profit distributed within  a given year or conference. Wins tend to cluster in the middle, with the best teams usually having the most wins. While this does not directly correlate because some teams who have a lot of wins are not in power conferences. A team could be well above 30 wins, however play in a terrible conference and have not played any other team capable of playing in the March Madness tournament. Revenue and profit for each team is also interesting. Power conference schools can easily spread the data a part. Most mid major teams do not operate with a large budget and are therefore all in a small area. While bigger schools have more revenue and possible profits but could also lose money because they are spending so much.


  The relationship plot adds another layer by showing how financial differences show up on the court. If you plot revenue of profit against wins, there is a clear trend between teams with more profit or revenue and more wins. There are also clear outliers in this data. As any one cinderella team would most likely have much less revenue than a power 5 team. Thus the reason for these teams normally do not do as well in the tournament. Basketball is highly influenced by many things and one team getting hot at the right time can lead to a deep run and everyone rooting for them. Some things like team chemistry, coaching strategy, and player development are nearly impossible to plot and see. That is where smaller teams have success and prove as a good team. Overall the shiny app is good representation that many things impact performance. Wins generally happen more if you spend or make more money. However, just like the things I just touched on, some things are unpredictable and that is the best part about March Madness. 
  
  This project provides a great exploration of how financial performance and on court success is definitely related. However, things that cannot be tracked also lead to improved performance and shows the wide landscape of college basketball. One limitation of the project is the small time period. I wish the graphics could have been over a wider time period, possibly the whole 21st century. The five years I had still showed great depth but some things needed to be explained thoughout because they only make sense if you touch on something from years outside the data (such as louisville). Some other variables could have been helpful to make the visuals stronger, or the shiny app better. Such as having strength of schedule or coaching quality. 

  If I had more time, I would have looked to add more variables that I just touched on. I would have also explored the option for more years. I do think in today's world with so many advanced metrics it would be cool to see how those can predict success as well. Or compare the strength of new metrics to those that we used to use for the last 100 years. Overall, this project was interesting in many ways and showed pretty much what I thought. However, some teams do make a lot of money and with the new world of NIL I'm sure these graphs numbers would be exponentially larger. 




