---
title: "Final Presentation"
format: 
  revealjs:
    embed-resources: true
editor: visual
---

## College Basketball Spending

- The data set I am using depicts college basketball teams from 2015 through 2019

- Some of the most important variables I will be using are:


```{r}
library(dplyr)
library(tidyverse)
library(readr)
 library(stringr)
library(fuzzyjoin)
library(plotly)

sports <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-03-29/sports.csv')

basketball_stat <- readr::read_csv("data/College Basketball Dataset-2019.csv")

basketball <- readr::read_csv("data/College Basketball Dataset.csv")

basketball <- basketball |> filter(YEAR %in% c("2015", "2016", "2017", "2018"))

```

```{r}
basketball_2015_2019 <- bind_rows(basketball, basketball_stat) |>

  mutate(YEAR = ifelse(is.na(YEAR), "2019", YEAR)) |>
  select(TEAM, CONF, G, W, POSTSEASON, SEED, YEAR) |>
  mutate(TEAM = TEAM |>
  str_to_lower())


  basketball_2015_2019$YEAR <- as.numeric(basketball_2015_2019$YEAR)

basketball_2015_2019
```

```{r}
basketball_money <- sports |> filter(sports == "Basketball") |>
  filter(classification_code == "1")|>
  select(year, unitid, institution_name, state_cd, classification_code, classification_name, sector_cd, sector_name, partic_men, partic_women, sum_partic_men, sum_partic_women, rev_men, rev_women, total_rev_menwomen, exp_men, exp_women, total_exp_menwomen, sports) |>
  rename(YEAR = year) |>
  rename(TEAM = institution_name) |>
  mutate(TEAM = TEAM |>
           str_to_lower() |>
           str_replace_all("university", "") |>
           str_replace_all("college", "") |>
           str_replace_all("campus", "") |>
           str_replace_all("of ", "") |>
           str_replace_all("at ", "") |>
           str_replace_all("main ","") |>
           str_replace_all("station ", "") |>
           str_replace_all("-", " ") |>
           str_replace_all("chapel hill", "") |>
           str_replace_all("ann arbor", "" ) |>
           str_replace_all("madison", "") |>
           str_replace_all("raleigh", "" ) |>
           str_squish())
```

```{r}
df <- stringdist_left_join(
  basketball_2015_2019, basketball_money,
  by = "TEAM",
  method = "osa",           # Optimal String Alignment
  max_dist = 1,             # Allow up to 2 changes (add, drop, swap)
  distance_col = "str_dist"
)

df <- df |> filter(YEAR.x == YEAR.y) |>
  rename(YEAR = YEAR.x,
         TEAM = TEAM.x) |>
  select(- YEAR.y,
         - TEAM.y) |>
  mutate(in_tournament = !is.na(POSTSEASON)) |>
mutate(profit_men = rev_men - exp_men)
```





## How does profits of college basketball programs compare to wins?

- Since the data is before NIL, I am unable to predict wins based on how much money each school is spending

- Instead, I am interested in how each team does based on how much profit they are bringing in

- Typically the larger schools have larger arenas with much larger fan engagement

- While some of these larger schools may not be a great program, therefore losing money because they spend to much each year

## Blue Bloods

```{r}
blue_bloods <- df |>
 filter(TEAM %in% c("duke", "north carolina", "kentucky", "kansas", "villanova", "connecticut", "ucla")) |>
  group_by(TEAM)

blue_bloods_plot <- ggplot(data = blue_bloods, aes(x = YEAR, y = profit_men, color = TEAM,
                               label = POSTSEASON)) +
  geom_point() +
  theme_minimal() +
  scale_color_viridis_d() +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 24) +
  labs(y = "profit")
```

```{r}
ggplotly(blue_bloods_plot, tooltip = "label")
```


## In the Tournament vs Not in Tournament

```{r}
tourney_vs_no_tourney <- df |> group_by(in_tournament)

tourney_plot <- ggplot(tourney_vs_no_tourney, aes(x = YEAR, y = profit_men,  color = in_tournament,
                                  label = TEAM)) +
  geom_point() +
  geom_smooth(se = FALSE) +
    scale_y_continuous(labels = scales::comma) +
  labs(y = "profit") +
  theme_minimal(base_size = 24)
```

```{r}
ggplotly(tourney_plot, tooltip = "label")
```



## is a new slide (whatever is directly after is the slide title)

Style of presentation

Data of introduction: show data set or specific variables

Question of interest

Then slides depicted graphs that you have made

## USE base size within graph to make scales larger

in theme (base_size = 24)
