---
title: "README"
format: gfm
editor: visual
---


## College Basketball Spending

```{r}
library(dplyr)
library(tidyverse)
library(readr)
 library(stringr)
library(fuzzyjoin)
library(plotly)
library(shiny)



sports <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-03-29/sports.csv')

basketball_stat <- readr::read_csv("data/College Basketball Dataset-2019.csv")

basketball <- readr::read_csv("data/College Basketball Dataset.csv")

basketball <- basketball |> filter(YEAR %in% c("2015", "2016", "2017", "2018"))

```

```{r}
basketball_2015_2019 <- bind_rows(basketball, basketball_stat) |>

  mutate(YEAR = ifelse(is.na(YEAR), "2019", YEAR)) |>
  select(TEAM, CONF, G, W, POSTSEASON, SEED, YEAR) |>
  mutate(TEAM = TEAM |>
  str_to_lower())


  basketball_2015_2019$YEAR <- as.numeric(basketball_2015_2019$YEAR)

basketball_2015_2019
```

```{r}
basketball_money <- sports |> filter(sports == "Basketball") |>
  filter(classification_code == "1")|>
  select(year, unitid, institution_name, state_cd, classification_code, classification_name, sector_cd, sector_name, partic_men, partic_women, sum_partic_men, sum_partic_women, rev_men, rev_women, total_rev_menwomen, exp_men, exp_women, total_exp_menwomen, sports) |>
  rename(YEAR = year) |>
  rename(TEAM = institution_name) |>
  mutate(TEAM = TEAM |>
           str_to_lower() |>
           str_replace_all("university", "") |>
           str_replace_all("college", "") |>
           str_replace_all("campus", "") |>
           str_replace_all("of ", "") |>
           str_replace_all("at ", "") |>
           str_replace_all("main ","") |>
           str_replace_all("station ", "") |>
           str_replace_all("-", " ") |>
           str_replace_all("chapel hill", "") |>
           str_replace_all("ann arbor", "" ) |>
           str_replace_all("madison", "") |>
           str_replace_all("raleigh", "" ) |>
           str_squish())
```

```{r}
df <- stringdist_left_join(
  basketball_2015_2019, basketball_money,
  by = "TEAM",
  method = "osa",           # Optimal String Alignment
  max_dist = 1,             # Allow up to 2 changes (add, drop, swap)
  distance_col = "str_dist"
)

df <- df |> filter(YEAR.x == YEAR.y) |>
  rename(YEAR = YEAR.x,
         TEAM = TEAM.x) |>
  select(- YEAR.y,
         - TEAM.y) |>
  mutate(in_tournament = !is.na(POSTSEASON)) |>
mutate(profit_men = rev_men - exp_men)
```

During my Data Visualization class, my final project focuses on college basketball spending from 2015 through 2019. The goal of my data is to determine if a team who makes more profit from their basketball teams ultimately leads to more wins. Throughout the project, I made a multitude of visuals to display this along with a shiny application. There is a slight trend that shows more profitable basketball teams correlates with more wins and a better finish in March Madness. With this being prior to NIL, the world of basketball has changed and the spending on college basketball teams has not only changed, but ballooned in size.



```{r}
tourney_vs_no_tourney <- df |> group_by(in_tournament)

tourney_plot <- ggplot(tourney_vs_no_tourney, aes(x = YEAR, y = profit_men,  color = in_tournament,
                                  label = TEAM)) +
  geom_point() +
  geom_smooth(se = FALSE) +
    scale_y_continuous(labels = scales::comma) +
  labs(title = "Analyzing Teams in the Tournament vs not in Tournament"  ,
       y = "profit")
```

```{r}
ggplotly(tourney_plot, tooltip = "label")
```

The visual above depicts teams who made the NCAA March Madness tournament from 2015 to 2019 with profit of each team during each year. Based on the visual, there is clear correlation between teams who made the tournament versus those who did not. Teams who made the tournament were in the range of 5 million dollars profit each year. While teams who did not make the tournament were in the realm of 500,000 dollars to maybe $1.5 million.


Profits versus Postseason success

```{r}
df <-  df |> mutate(POSTSEASON = as.numeric(factor(POSTSEASON,
    levels = c("R68", "R64", "R32", 
               "S16", "E8", "F4", "2ND" ,"Champions")
  )))
```

```{r}
finish_vs_profit <- ggplot(df, aes(x = profit_men, y = POSTSEASON, color = W,
                                   label = TEAM)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Team Profit vs. Postseason Success",
    x = "Profit",
    y = "Postseason Round"
  ) +
    scale_color_viridis_c() +
  scale_x_continuous(labels = scales::comma)
```

```{r}
ggplotly(finish_vs_profit, tooltip = "label")
```



  This visual does a good job showing a different view from just direct profit versus postseason success. Although there is definitely evidence of this. Regular season wins have a huge effect as well. There have been only 5 teams in the 21st century to win the tournament as 3 seed or lower. Three of those being three seeds, one four seed, and one seven seed. To win March Madness, the overwhelming majority says that you must be high seed to achieve great success in the tournament. 


